A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-22 tex/body.tex
minor

By Theorems~\ref{thm:t},~\ref{thm:rs}, and~\ref{thm:ss}, the
calculus is sound and complete for the ground $\thD$ theory, and we can
rightly call it a decision procedure.

@@

The ground decision procedure for (co)datatypes
was presented at a high level of abstraction. This section describes
the main aspects of the implementation within the SMT solver CVC4:\ the
integration into CDCL($T$), the extension to quantified formulas, and
some of the optimizations.

@@

The efficiency properties of the procedure are mostly as described in Barrett et
al.\ \cite{barrett-et-al-2007}. In particular, it is theoretically and
empirically advantageous to delay applications of \rn{Split} as long as
possible. Similarly, \rn{Acyclic} and \rn{Unique} are fairly expensive because
they require traversing the equivalence classes, which is why they are part of
phase~2.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories by Peter Gammie and Andreas Lochbihler (\gandl, 317
goals), about Bird and Stern--Brocot trees. To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all three benchmark suites. It accounts for an overall
success rate increase of over \OK{5\%}, which is quite significant. For Isabelle
users, the practical usefulness of interpreted (co)datatypes appears roughly
comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}. The raw data
also reveals that the stronger decision procedures almost fully subsume the weaker
ones.

>>>>

By Theorems~\ref{thm:t},~\ref{thm:rs}, and~\ref{thm:ss}, the
calculus is sound and complete for the ground $\thD$ theory. We can
rightly call it a decision procedure.

@@

The ground decision procedure for (co)datatypes
was presented at a high level of abstraction, omitting quite a few details. This
section describes the main aspects of the implementation within the SMT solver
CVC4:\ the integration into CDCL($T$), the extension to quantified formulas, and
some of the optimizations.

@@

The efficiency properties of the procedure are mostly as described by Barrett et
al.\ \cite{barrett-et-al-2007}. In particular, it is both theoretically and
empirically beneficial to delay applications of \rn{Split} as long as
possible. Similarly, \rn{Acyclic} and \rn{Unique} are fairly expensive because
they require traversing the equivalence classes, which is why they are part of
phase~2.

@@

To evaluate the decision procedure, we generated benchmark problems from existing
Isabelle formalizations using Sledgehammer.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the theory files from the Isabelle distribution (Distro, 879~goals)
and the \emph{Archive of Formal Proofs} (AFP, 2974~goals) \cite{klein-et-al-afp}
that define codatatypes falling within the supported fragment. We added
two unpublished theories about Bird and Stern--Brocot trees
by Peter Gammie and Andreas Lochbihler (\gandl,
317~goals). To exercise the datatype
support, the benchmarks were complemented by theories
about lists and trees. The theories were selected before running the
experiments. The experimental data are publicly available.%
\footnote{\url{http://lara.epfl.ch/~reynolds/CADE2015-cdt/}}
%http://www21.in.tum.de/~blanchet/codp_data.tgz}}

@@

The results are summarized in Table~\ref{tab:bench}. The decision procedure
makes a difference across all three benchmark suites. It accounts for an overall
success rate increase of over \OK{5\%}, which is quite significant. For Isabelle
users, the practical usefulness of interpreted (co)datatypes is roughly
comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
obviously depends a lot on the application. The raw evaluation data also
reveals that the stronger decision procedures almost completely subsume the
weaker ones in practice.
